\textls[-0]{\thetitle}

### 2023-03-15 canonicitymodel.tex
Save Canonicity model.

We can roughly consider each Grothendieck Universe $\Set{i}$ as a $\Set{i}$ in Agda. In other words, each $\Set{i}$ is 
\begin{itemize}
  \item Closed under dependent function type and dependent tuple type. For example, later, for our interpretation of dependent function type, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have $\denotesC{\TyPi{A}{B}} \in \Set{i}$
  \item the hierarchy is cumulative as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$
\end{itemize}

@@

\begin{align*}
  \denotesC{\goodCtx{\Gm}{k}} \text{ is a function }&: \{\gm : \goodSub{\cdot}{\gm}{\Gm} \} \to \Set{k} \quad \text{ (also sets indexed by closed substitution)}\\
  % &\text{and we define } \goodCtxS{\Gm}{i} :\iff \denotesC{\goodCtx{\Gm}{i}} \text{ is defined} \\
  \denotesC{\goodType{\Gm}{T}{j}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}}\prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\{t : \goodTerm{\cdot}{t}{\sub{T}{\gm}}\}\to\Set{j} \\
  % &\text{and we define } \goodTypeS{\Gm}{T}{i} : \iff \denotesC{\goodType{\Gm}{T}{i}} \text{ is defined } \\ 
  \denotesC{\goodSub{\Gm}{\delta}{\Delta}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)} \denotesC{\goodCtx{\Dl}{}}(\delta \circ \gm) \\
  % &\text{and we define } \goodSubC{\Gm}{\gm}{\Delta} :\iff \denotesC{\goodSub{\Gm}{\gm}{\Delta}} \text{ is defined} \\
  \denotesC{\goodTerm{\Gm}{t}{T}} \text{ is a dependent function}&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})\\
  \denotesC{\goodType{\Gm}{\top}{}}(\gm)(\gm')(t) &= \{\star \} \\
  \denotesC{\goodType{\Gm}{\bot}{}}(\gm)(\gm')(t) &= \emptyset \\
  \denotesC{\goodType{\Gm}{\cB}{}}(\gm)(\gm')(t) &= \{\star^1 \ | \text{ when } t \equiv \true  \} \dot{\cup} \{\star^2 \ | \text{ when } t \equiv \false  \} \quad \text{disjoint union} \\
  \denotesC{\goodType{\Gm}{\TyId{a}{b}}{}}(\gm)(\gm')(t) &= \{\star \ |  \text{ when } t \equiv \eqrefl{\sub{a}{\gm}} \text{ and } \sub{a}{\gm} \equiv \sub{b}{\gm} \} \\
  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})((\gm', u'))(\sub{\app{t}}{\SubstExt{"id"}{u}})\\
  \denotesC{\goodType{\Gm}{\TySigma{A}{B}}{}}(\gm)(\gm')(t) &= \sum_{u' \in \denotesC{A}(\gm)(\gm')} \\
  \denotesC{\goodCtx{\cdot}{}}(\gm) &= \{\star\} \quad \text{  a singleton set } \\ 
  \denotesC{\goodCtx{\Gm, T}{}}(\gm_t) &= \{ (\gamma , t) : \gamma \in \denotesC{\goodCtx{\Gm}{}}(\pi_1 \gm_t), t \in \denotesC{\goodType{\cdot}{\sub{T}{\gm}}{}}(\pi_2 \gm_t)  \} \\
  \denotesC{\goodTerm{\Gm}{\unit}{\top}}(\gm)(\gm') &= \star\\
  \denotesC{\goodTerm{\Gm}{\true}{\cB}}(\gm)(\gm') &= \star^1 \\
  \denotesC{\goodTerm{\Gm}{\false}{\cB}}(\gm)(\gm') &= \star^2 \\
  \denotesC{\goodTerm{\Gm}{\ifb{c}{a}{b}}{T}}(\gm)(\gm') &= \denotesC{a}(\gm) \text{ if } \denotes{c}(\gm)(\gm') = \star^1 \\
  \denotesC{\goodTerm{\Gm}{\ifb{c}{a}{b}}{T}}(\gm)(\gm') &= \denotesC{b}(\gm)(\gm') \text{ if } \denotes{c}(\gm)(\gm') = \star^2 \\
  \denotesC{\goodTerm{\Gm}{\eqrefl{t}}{\TyId{t}{t}}}(\gm)(\gm') &= \star\\
  \denotesC{\goodTerm{\Gm}{\lam{t}}{\TyPi{A}{B}}} &= \\
  \denotesC{\goodTerm{\Gm}{\app{t}}{B}} &= \\
  \denotesC{\goodTerm{\Gm}{\pair{a}{b}}{\TySigma{A}{B}}} &= \\
  \denotesC{\goodTerm{\Gm}{\fst{t}}{T}} &= \\
  \denotesC{\goodTerm{\Gm}{\snd{t}}{T}} &= \\
  \denotesC{\goodType{\Gm}{\cU_j}{j+1}}(\gm)(\gm')(T) &= \{t : \goodTerm{\cdot}{t}{\El{T}}\}  \to \Set{j} \\
  \denotesC{\goodTerm{\Gm}{\codety{T}}{\cU_j}}(\gm)(\gm') &= \denotesC{T}(\gm)(\gm') \\ 
  \denotesC{\goodType{\Gm}{\El{T}}{j}}(\gm)(\gm')(t) &= \denotesC{T}(\gm)(\gm')(t) \\
\end{align*}

@@

The counter-intuitive part of the proof-relevant logical relation is the treatment of proposition---a proposition is encoded as a subsingleton data. For example, $\denotesC{\goodType{\Gm}{\top}{}} = $ $ \{\star \ | \text{  when } t \equiv \unit \}$ represents a singleton set when $t \equiv \unit$ or an empty set otherwise. However, not all the denotation of types are subsingleton---the salient counter-examples are $\denotes{\goodType{\Gm}{\cU}{}}$ and $\denotes{\goodType{\Gm}{\TyPi{A}{B}}{}}$. The non-trivial interpretation on these types is the main reason we use proof-relevant logical relations.

>>>>

We can roughly consider each Grothendieck Universe $\Set{i}$ as a $\Set{i}$ in Agda. In other words, each $\Set{i}$ is 
\begin{itemize}
  \item Closed under dependent function type and dependent tuple type. For example, later, for our interpretation of dependent function type, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have $\denotesC{\TyPi{A}{B}} \in \Set{i}$
  \item the hierarchy is cumulative as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$
  \item Thus if $\denotesC{A} \in \Set{i}, \denotesC{B} \in \Set{j}$, we will have $\denotesC{\TyPi{A}{B}} \in \Set{i} \cup \Set{j} = \Set{i \lcup j}$
\end{itemize}

@@

\begin{align*}
  \denotesC{\goodCtx{\Gm}{k}} \text{ is a function }&: \{\gm : \goodSub{\cdot}{\gm}{\Gm} \} \to \Set{k} \quad \text{ (also sets indexed by closed substitution)}\\
  % &\text{and we define } \goodCtxS{\Gm}{i} :\iff \denotesC{\goodCtx{\Gm}{i}} \text{ is defined} \\
  \denotesC{\goodType{\Gm}{T}{j}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}}\prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\{t : \goodTerm{\cdot}{t}{\sub{T}{\gm}}\}\to\Set{j} \\
  % &\text{and we define } \goodTypeS{\Gm}{T}{i} : \iff \denotesC{\goodType{\Gm}{T}{i}} \text{ is defined } \\ 
  \denotesC{\goodSub{\Gm}{\delta}{\Delta}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)} \denotesC{\goodCtx{\Dl}{}}(\delta \circ \gm) \\
  % &\text{and we define } \goodSubC{\Gm}{\gm}{\Delta} :\iff \denotesC{\goodSub{\Gm}{\gm}{\Delta}} \text{ is defined} \\
  \denotesC{\goodTerm{\Gm}{t}{T}} \text{ is a dependent function}&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})\\
  \denotesC{\goodType{\Gm}{\top}{}}(\gm)(\gm')(t) &= \{\star \} \\
  \denotesC{\goodType{\Gm}{\bot}{}}(\gm)(\gm')(t) &= \emptyset \\
  \denotesC{\goodType{\Gm}{\cB}{}}(\gm)(\gm')(t) &= \{\star^1 \ | \text{ when } t \equiv \true  \} \dot{\cup} \{\star^2 \ | \text{ when } t \equiv \false  \} \quad \text{disjoint union} \\
  \denotesC{\goodType{\Gm}{\TyId{a}{b}}{}}(\gm)(\gm')(t) &= \{\star \ |  \text{ when } t \equiv \eqrefl{\sub{a}{\gm}} \text{ and } \sub{a}{\gm} \equiv \sub{b}{\gm} \} \\
  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})((\gm', u'))(\sub{\app{t}}{\SubstExt{"id"}{u}})\\
  \denotesC{\goodType{\Gm}{\TySigma{A}{B}}{}}(\gm)(\gm')(t) &= \sum_{u' \in \denotesC{A}(\gm)(\gm')(\fst{t})} \denotesC{B}(\SubstExt{\gm}{\fst{t}})((\gm', u'))(\snd{t}) \\
  \denotesC{\goodCtx{\cdot}{}}(\gm) &= \{\star\} \quad \text{  a singleton set } \\ 
  \denotesC{\goodCtx{\Gm, T}{}}(\gm_t) &= \{ (\gamma , t) : \gamma \in \denotesC{\goodCtx{\Gm}{}}(\pi_1 \gm_t), t \in \denotesC{\goodType{\cdot}{\sub{T}{\gm}}{}}(\pi_2 \gm_t)  \} \\
  \denotesC{\goodTerm{\Gm}{\unit}{\top}}(\gm)(\gm') &= \star\\
  \denotesC{\goodTerm{\Gm}{\true}{\cB}}(\gm)(\gm') &= \star^1 \\
  \denotesC{\goodTerm{\Gm}{\false}{\cB}}(\gm)(\gm') &= \star^2 \\
  \denotesC{\goodTerm{\Gm}{\ifb{c}{a}{b}}{T}}(\gm)(\gm') &= \denotesC{a}(\gm)(\gm') \quad \text{ if } \denotes{c}(\gm)(\gm') = \star^1 \\
  \denotesC{\goodTerm{\Gm}{\ifb{c}{a}{b}}{T}}(\gm)(\gm') &= \denotesC{b}(\gm)(\gm') \quad \text{ if } \denotes{c}(\gm)(\gm') = \star^2 \\
  \denotesC{\goodTerm{\Gm}{\eqrefl{t}}{\TyId{t}{t}}}(\gm)(\gm') &= \star\\
  \denotesC{\goodTerm{\Gm}{\lam{t}}{\TyPi{A}{B}}}(\gm)(\gm') &= \lambda u \lambda u' . \denotesC{t}(\SubstExt{\gm}{u}) ((\gm' , u')) \\
  \denotesC{\goodTerm{\Gm}{\app{t}}{B}}(\gm)(\gm') &= \denotesC{t}(\pi_1 \gm)(\gm'[0])(\pi_2 \gm)(\gm'[1]) \\
  \denotesC{\goodTerm{\Gm}{\pair{a}{b}}{\TySigma{A}{B}}}(\gm)(\gm') &= (\denotesC{a}(\gm)(\gm') , \denotesC{b}(\gm)(\gm') )\\
  \denotesC{\goodTerm{\Gm}{\fst{t}}{T}}(\gm)(\gm') &= \denotesC{t}(\gm)(\gm')[0] \\
  \denotesC{\goodTerm{\Gm}{\snd{t}}{T}}(\gm)(\gm') &= \denotesC{t}(\gm)(\gm')[1]\\
  \denotesC{\goodType{\Gm}{\cU_j}{j+1}}(\gm)(\gm')(T) &= \{t : \goodTerm{\cdot}{t}{\El{T}}\}  \to \Set{j} \\
  \denotesC{\goodTerm{\Gm}{\codety{T}}{\cU_j}}(\gm)(\gm') &= \denotesC{T}(\gm)(\gm') \\ 
  \denotesC{\goodType{\Gm}{\El{T}}{j}}(\gm)(\gm')(t) &= \denotesC{T}(\gm)(\gm')(t) \\
\end{align*}

@@

The counter-intuitive part of the proof-relevant logical relation is the treatment of proposition---a proposition is encoded as a subsingleton data. For example, $\{\star^1 \ | \text{ when } t \equiv \true  \} \dot{\cup} \{\star^2 \ | \text{ when } t \equiv \false  \}$ represents a disjoint union of two subsingleton set. The former one will be a singleton if we have a proof $t \equiv \true$ or an empty set otherwise, same as the later one.  In other words, $\star^1$ witness a proof that $t \equiv true$. The same idea applies to the latter set.
